A Switching Planner for Combined Task and Observation Planning

### 2010-11-26 code/subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/dtpddl.tex
New notations, and corrections.

The syntax and semantics for describing initial state distributions in
DTPDDL is taken verbatim from PPDDL. It is useful to present that
factored tree-like structure here, in order to aid us when discussing
our switching continual planner. The distribution is expressed in a
tree-like structure of terms. Each term is either: (1) atomic, e.g., a
state proposition such as $(=(\pp{is-in}~\pp{box})~\pp{office})$,
(2) probabilistic, e.g., $(\pp{probabilistic}~p_1 (T_1) .. p_n (T_n))$
where $T_i$ are conjunctive, or (3) a conjunct over probabilistic and
atomic terms. The root term is always conjunctive, and the leaves are
atomic. For example, the starting distribution for $\pp{DORA}$ might
be described thusly:\footnote{In PDDL, $(\pp{:init}~T_1..T_n)$
expresses the conjunctive root of the tree -- i.e., the root node
$(\pp{and}~T_1..T_n)$. Also, we shall write $\prop$, rather than
$(\pp{and}~\prop)$, it a conjunctive term contain a single atomic
subterm.}

@@

\noindent The interpretation of such an expression can be given
according to a {\em visitation} of terms. An atom is {\em visited} iff
its conjunctive parent is visited, and a conjunctive term is visited
iff all its immediate subterms are visited. A probabilistic term is
visited iff its conjunctive parent is visited, and exactly one of its
subterms, $T_i$, is visited. Each visitation of the root term
according to this recursive definition encapsulates a starting state,
along with the probability it occurs. The former corresponds to the
union of all visited atoms, and the latter corresponds to the product
of $p_i$ entries on the visited subterms of probabilistic
elements. Making these ideas concrete, our example yields the
following flat state distribution:

>>>>

The syntax and semantics for describing initial state distributions in
DTPDDL is taken verbatim from PPDDL. It is useful to present that
factored tree-like structure here, in order to aid us when discussing
our switching continual planner. The distribution is expressed in a
tree-like structure of terms. Each term is either: (1) atomic, e.g., a
state proposition such as $(=(\pp{is-in}~\pp{box})~\pp{office})$, (2)
probabilistic, e.g., $(\pp{probabilistic}~\prob_1 (T_1) .. \prob_n
(T_n))$ where $T_i$ are conjunctive, or (3) a conjunct over
probabilistic and atomic terms. The root term is always conjunctive,
and the leaves are atomic. For example, the starting distribution for
$\pp{DORA}$ might be described thusly:\footnote{In PDDL,
$(\pp{:init}~T_1..T_n)$ expresses the conjunctive root of the tree --
i.e., the root node $(\pp{and}~T_1..T_n)$. Also, we shall write
$\prop$, rather than $(\pp{and}~\prop)$, it a conjunctive term
contains a single atomic subterm.}

@@

\noindent The interpretation of such an expression can be given
according to a {\em visitation} of terms. An atom is {\em visited} iff
its conjunctive parent is visited, and a conjunctive term is visited
iff all its immediate subterms are visited. A probabilistic term is
visited iff its conjunctive parent is visited, and exactly one of its
subterms, $T_i$, is visited. Each visitation of the root term
according to this recursive definition encapsulates a starting state,
along with the probability it occurs. The former corresponds to the
union of all visited atoms, and the latter corresponds to the product
of $\prob_i$ entries on the visited subterms of probabilistic
elements. Making these ideas concrete, our example yields the
following flat state distribution:

### 2010-11-26 code/subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/switching.tex
New notations, and corrections.

A sequential session uses a classical planner to compute a trace. The
latter encapsulates assumptions about: (1) the true underlying state,
(2) how the plan's execution will progress, and (3) the possibility of
the agent eventually holding strong beliefs about the truth values of
specific state propositions. Here we describe the deterministic
planning problem, derived from the DTPDDL model, that admits plans
which correspond to traces.

@@

\noindent In detail, a valid assumption in our setting corresponds to a {\em
relaxed} visitation of the root term of $(\pp{:init})$. In this
relaxed case, a conjunctive term is visited iff its atomic subterms
are visited, and zero or more of its immediate probabilistic subterms
are visited. The starting state, $\state_0$, for sequential planning
is an {\em abstract} state that captures the unique assumption which
has probability $1.0$. In $\pp{DORA}$, that is:

@@

In order to make assumptions available to the sequential planner, we
add one {\em assumptive} action $\assumptiveS{i}$ to the problem for
each element, $p_i (T_i)$, of each probabilistic term from
$(\pp{:init})$. The physics of these actions is as
follows. $\assumptiveS{i}$ can be executed if no $\assumptiveS{j}$,
$j \neq i$, has been executed from the same probabilistic term, and,
either $(\pp{probabilistic}~..p_i~(T_i)..)$ is in the root conjunct,
or it occurs in $T_k$ for some executed $\assumptiveS{k}$.
%%
Executing $\assumptiveS{i}$ in $s$ effects a transition to a successor
state $s^{T_i}$ with probability $p_i$, and $s^\bot$ with probability
$1 - p_i$. Here, $s^{T_i}$ is the union of $s$ with atomic terms from
$T_i$ and conflicting propositions from $s$ deleted. For example, if
the action adds $(=(\pp{is-in}~\pp{box})~\pp{office})$, then it must
delete $(=(\pp{is-in}~\pp{box})~\#)$. State $s^\bot$ is an added sink,
where $\forall\action\in\actions$ $\reward(\state^\bot, \action) = 0$
and $\transProb(\state^\bot, \action, \state^\bot) = 1$.

For sequential planning the operators from the POMDP model are
available with semantics that accommodate the above abstraction as
follows. In the deterministic model a proposition $\prop$ can be
thought to have a ternary interpretation at a state \state, as either
{\em true}, written $\prop\in\state$, {\em false}, written
$\prop\not\in\state$, or {\em unspecified}, with a slight abuse of
notation written $\prop\#\in\state$. For example, in $\pp{DORA}$
$(=(\pp{is-in}~\pp{box})\pp{office})\#\in\state_0$. For POMDP actions
$\action\in\actions$ in the deterministic model, if
$\prop \in \poss(\action)$, or if $\prop$ is the subject of a positive
or negative effect of $\action$, then $\action$ is not applicable
in \state\ if $\prop\#\in\state$. Indeed, first an assumptive action
must be applied to make a truth assignment to \prop, and then that
truth assignment could be modified by POMDP operators.

Addressing now the switching semantics of action execution. If
$\action\in\actions$ has a precondition that is not true with
probability $1$ in the underlying belief $\bstate$, then before it is
executed blindly, the system switches to a contingent session,
discussed in the following section. In detail, we halt the sequential
session at \action\ if $\exists
\state$ s.t. $\bstate(\state)>0$, 
$\prop\in\poss(\action)$ , and $\prop\not\in\state$. According to the
semantics of action execution there must be a $\state'$ s.t.
$\bstate(\state')>0$ and $\prop\in\state'$. Otherwise, \action\ must
not have been scheduled at \bstate.

then $\reward(\state,\action)=10$, and
otherwise if $\prop\not\in\state'$ then
$\reward(\state',\action)=-10$.

@@

\noindent which is equal to the maximal additive contribution a trace of the plan can
make to Eq~\ref{eq:expectedvalue}, taking an FSC that corresponds to
the sequential plan.  In goal directed problems, where the only
non-zero reward is received at the first transition to a goal state,
the Eq~\ref{eq:tracevalue} criteria gives us the behaviour
of \system{FFR$_a$}~\cite{yoon:etal:2007}. From the perspective of
gradient-based reinforcement learning systems, such as William's
REINFORCE and GPOMDP variants for planning~\cite{olivier:doug:2009},
under reasonable assumptions the optimal trace identifies a maximal
gradient step from a uniformly random soft-max policy. Finally, it is
worth clarifying that although the above discussion has been in terms
of a trace of length $N$, there is nothing in our approach that
artificially limits the length of candidate sequential plans.

@@

In a contingent session our switching planner solves a sensing problem
in an abstract processes defined in terms of the assumptive actions in
the current trace, and the $(\pp{:init})$ declaration from the
original problem description.

@@

\noindent Where $\$(T_i)$ is a positive numeric quantity which
captures the reward the agent receives for correctly rejecting a
false assumption. The added dual actions are propositionally trivial
-- i.e., $\forall\state\in\states$ $\poss(\assumptiveDT{i})
\subseteq \state$ and $\assumptiveDT{i}(\state) \equiv \state$.  
%%
In order to reduce the number of possible rejections we do not
consider assumptions that are not {\em active} with respect to the
action, $\action_i$, whose scheduled execution switch the system into
the contingent session. A assumption $\assumptiveS{i}$ is active if
$\action_i$ is not applicable in the sequential model unless
$\assumptiveS{i}$ is scheduled in the trace prefix to $\action_i$. For
example, if $\pp{DORA}$'s trace is:

@@

\noindent Taking the switching action $\action_i$ to be
$(\pp{look-for-object}~\pp{DORA}~\pp{box}~\pp{office})$, we have that
$\actions^{\circ}(.3;(=(\pp{is-in}~\pp{cup})\pp{kitchen}))$ is not
active, and therefore exclude
$\actions^{\bullet}(.3;(=(\pp{is-in}~\pp{cup})\pp{kitchen}))$ from the
POMDP posed to the contingent session.

Again taking $\action_i$ to be the switching action, we include
another propositionally trivial action $\actions.\poss(\action_i)$
with the reward property:

\[
\reward(\state, \actions.\poss(\action_i)) = \bigg\{ \begin{array}{ll}
\$(\poss(\action_i)) & \pp{if}~\;\; \poss(\action_i) \subseteq \state \\
-\$(\poss(\action_i)) & \pp{otherwise} \\
\end{array}
\]

For the purposes of successive switching, execution of either a dual
action of the form $\assumptiveDT{i}$, or the
$\actions.\poss(\action_i)$, returns control to a sequential
session. If a dual is executed, then the sequential planner must
replan with $\bstate_0$ equal to the underlying
belief-state. Otherwise, if $\actions.\poss(\action_i)$ is executed,
then the current sequential plan is executed until further sensing is
scheduled, or to completion. In our project the rewards, i.e.,
function $\$:2^\propositions\to\Re$, for the POMDP posed to the
contingent session are sourced from a motivational component of our
robotic architecture.

@@

\noindent At this point, the session would proceed in an
abstraction of the environment that does not contain a
$\pp{cup}$. With regards to abstraction, the contingent and serial
sessions apply the same applicability conditions ---i.e., that each
effect and condition $\prop$ is specified $\prop\in\state$
(resp. $\prop\#\in\state$)--- to actions from the DTPDDL problem
description.

>>>>

A sequential session uses a classical planner to compute a trace. The
latter encapsulates assumptions about: (1) the true underlying state,
(2) how execution will progress, and (3) the possibility of the agent
eventually holding strong beliefs about the truth values of specific
state propositions. Here we describe the deterministic planning
problem, derived from the DTPDDL model, that admits plans which
correspond to traces.

@@

\noindent An assumption corresponds to a {\em
relaxed} visitation of the root term of $(\pp{:init})$. In this
relaxed case, a conjunctive term is visited iff its atomic subterms
are visited, and zero or more of its immediate probabilistic subterms
are visited. The starting state, $\state_0$, for sequential planning
is an {\em abstract} state that captures the unique assumption which
has probability $1.0$. In $\pp{DORA}$, that is:

@@

In order to make assumptions available to the sequential planner, we
add one {\em assumptive} action $\assumptiveS{i}$ to the problem for
each element, $\prob_i (T_i)$, of each probabilistic term from
$(\pp{:init})$. The physics of these actions is as
follows. $\assumptiveS{i}$ can be executed if no $\assumptiveS{j}$,
$j \neq i$, has been executed from the same probabilistic term, and,
either $(\pp{probabilistic}~..\prob_i~(T_i)..)$ is in the root conjunct,
or it occurs in $T_k$ for some executed $\assumptiveS{k}$.
%%
Executing $\assumptiveS{i}$ in $\state$ effects a transition to a
successor state $\state^{T_i}$ with probability $\prob_i$, and
$\state^\bot$ with probability $1 - \prob_i$. Here, $\state^{T_i}$ is
the union of $\state$ with atomic terms from $T_i$ and conflicting
propositions from $s$ deleted. For example, if the action adds
$(=(\pp{is-in}~\pp{box})~\pp{office})$, then it must delete
$(=(\pp{is-in}~\pp{box})~\#)$. State $\state^\bot$ is an added sink,
where $\forall\action\in\actions$ $\reward(\state^\bot, \action) = 0$
and $\transProb(\state^\bot, \action, \state^\bot) = 1$.

For sequential planning the operators from the POMDP model are
available with semantics that accommodate the above abstraction as
follows. In the deterministic model a proposition $\prop$ can be
thought to have a ternary interpretation at a state \state, as either
{\em true}, written $\prop\in\state$, {\em false}, written
$\prop\not\in\state$, or {\em unspecified}, with a slight abuse of
notation written $\prop\#\in\state$. For example, in $\pp{DORA}$
$(=(\pp{is-in}~\pp{box})\pp{office})\#\in\state_0$. For POMDP actions
$\action\in\actions$ in the deterministic model, if
$\prop \in \poss(\action)$, or if $\prop$ is the subject of a positive
or negative effect of $\action$, then $\action$ is not applicable
in \state\ if $\prop\#\in\state$. In other words, an assumptive action
must be applied to make a truth assignment to \prop, and then that
truth assignment could be modified by the DTPDDL domain operators.
%%
Lastly, With regards to switching, we have that if $\action$ has a
precondition that is not true with probability $1$ in the underlying
belief $\bstate$, then immediately before it is executed blindly, the
system switches to a contingent session, discussed in the following
section.

%% Addressing now the switching semantics of action execution. I
%% %% In detail, we halt the sequential
%% session at \action\ if $\exists
%% \state$ s.t. $\bstate(\state)>0$, 
%% $\prop\in\poss(\action)$ , and $\prop\not\in\state$. According to the
%% semantics of action execution there must be a $\state'$ s.t.
%% $\bstate(\state')>0$ and $\prop\in\state'$. Otherwise, \action\ must
%% not have been scheduled at \bstate.

%% then $\reward(\state,\action)=10$, and
%% otherwise if $\prop\not\in\state'$ then
%% $\reward(\state',\action)=-10$.

@@

\noindent which is equal to the maximal 
additive contribution a trace of the plan can make to
Eq~\ref{eq:expectedvalue}.  In goal directed problems, where the only
non-zero reward is received at the first transition to a goal state,
the Eq~\ref{eq:tracevalue} criteria gives us the behaviour
of \system{FFR$_a$}~\cite{yoon:etal:2007}. From the perspective of
gradient-based reinforcement learning systems, such as William's
REINFORCE and GPOMDP variants for planning~\cite{olivier:doug:2009},
under reasonable assumptions the optimal trace identifies a maximal
gradient step from a uniformly random soft-max policy. Finally, it is
worth clarifying that although the above discussion has been in terms
of a trace of length $N$, there is nothing in our approach that
artificially limits the length of candidate sequential plans.

@@

In a contingent session our switching planner solves a sensing problem
in an abstract processes defined in terms of the assumptive actions in
the current trace, and the $(\pp{:init})$ declaration that models the
underlying belief-state.

@@

\noindent Where $\$(T_i)$ is a positive numeric quantity which
captures the reward the agent receives for correctly rejecting a
false assumption. The added dual actions are propositionally trivial
-- i.e., $\forall\state\in\states$ $\poss(\assumptiveDT{i})
\subseteq \state$ and $\assumptiveDT{i}(\state) \equiv \state$.  
%%
In order to reduce the number of possible rejections we omit
assumptions that are not {\em active} with respect to the action,
\switchAction, whose scheduled execution switched the system into
the contingent session. An assumption $\assumptiveS{i}$ is active if
\switchAction\ is not applicable in the sequential model unless
$\assumptiveS{i}$ is scheduled in the trace prefix
to \switchAction. For example, if $\pp{DORA}$'s trace is:

@@

\noindent Taking the switching action \switchAction\ to be
$(\pp{look-for-object}~\pp{DORA}~\pp{box}~\pp{office})$, we have that
$\actions^{\circ}(.3;(=(\pp{is-in}~\pp{cup})\pp{kitchen}))$ is not
active, and therefore exclude
$\actions^{\bullet}(.3;(=(\pp{is-in}~\pp{cup})\pp{kitchen}))$ from the
POMDP posed to the contingent session.

We also include another propositionally trivial action
$\actions.\poss(\switchAction)$ with the reward property:

\[
\reward(\state, \actions.\poss(\switchAction)) = \bigg\{ \begin{array}{ll}
\$(\poss(\switchAction)) & \pp{if}~\;\; \poss(\switchAction) \subseteq \state \\
-\$(\poss(\switchAction)) & \pp{otherwise} \\
\end{array}
\]

For the purposes of successive switching, execution of either a dual
action of the form $\assumptiveDT{i}$, or the
$\actions.\poss(\switchAction)$, returns control to a sequential
session. If a dual is executed, then the sequential planner must
replan with $\bstate_0$ equal to the underlying
belief-state. Otherwise, if $\actions.\poss(\switchAction)$ is
executed, then the current sequential plan is executed until further
sensing is scheduled, or to completion. Lastly, we note that in our
project the rewards, i.e., function $\$:2^\propositions\to\Re$, for
the POMDP posed to the contingent session are sourced from a
motivational component of our robotic architecture. We make no
assumptions about how they are derived.

@@

\noindent At this point, the session would proceed in an
abstraction of the environment that does not contain a
$\pp{cup}$. With regards to abstraction, the contingent and serial
sessions apply the same applicability conditions ---i.e., that each
effect and condition $\prop$ is specified, $\prop\in\state$--- to
actions from the DTPDDL problem description.
